Hilbert space approximate Bayesian Gaussian processes using Stan

### 2020-03-25 Paper/manuscript_v2/manuscript_07_01.tex
fix last Michael's comments

\author{Gabriel Riutort-Mayol$^{1*}$, Michael R. Andersen$^2$, Paul-Christian Bürkner$^{3}$, Aki Vehtari$^{2}$}

\date{ \small
$^1$ Department of Cartographic Engineering, Geodesy, and Photogrammetry, Universitat Polit\`ecnica de Val\`encia, Spain. \break
$^2$ Department of Applied Mathematics and Computer Science, Technical University of Denmark, Denmark
$^3$ Department of Computer Science, Aalto University, Finland\break
$^*$ Corresponding author, Email: gabriuma@gmail.com
}

@@

The key element of a GP is the covariance function that defines the dependence structure between function values at different inputs. However, evaluating the covariance function \todo{evaluating the covariance function is not a problem, but computing the posterior is} comes with a computational issue because of the need of inverting its Gram matrix to optimize the hyperparameters. Given $n$ observations in the data, the computational complexity and memory requirements of computing the posterior distribution for a GP in general scale as $O(n^3)$ and $O(n^2)$, respectively. This limits their application to rather small data sets of a few tens of thousands observations at most. The problem becomes more severe when performing full Bayesian inference via sampling methods, where in each sampling step we need $O(n^3)$ computations when inverting the Gram matrix of the covariance function, usually through Cholesky factorization. To alleviate these computational demands, several approximate methods have been proposed.

@@

If $p(y_n|f(\bm{x}_n),\phi)=\mathcal{N}(y_n|f(\bm{x}_n),\sigma)$ then $\bm{f}$ can be integrated out analytically (with a computational cost of $O(n^3)$ for exact GPs and $O(nm^2)$ for sparse GPs). If $p(y_n|f(\bm{x}_n),\phi)=\mathcal{N}(y_n|f(\bm{x}_n),g(\bm{x}_n))$ \todo{Is something missing here? The function g is missing on the left hand side. } or $p(y_n|f(\bm{x}_n),\phi)$ is non-Gaussian, the marginalization does not have a closed form solution. Furthermore, if a prior distribution is imposed on $\phi$ and $\bm{\theta}$ to form a joint posterior for $\phi$, $\bm{\theta}$ and $\bm{f}$, approximate inference such as Markov chain Monte Carlo (MCMC) \citep{brooks_2011}, Laplace approximation (\citep{williams1998bayesian,rasmussen2006gaussian}, expectation propagation \citep{minka2001expectation}, or variational Bayes methods \citep{gibbs2000variational,csato2000efficient} are required. In this paper, we focus on the use of MCMC for integrating over the joint
posterior. MCMC is usually not the fastest approach, but it is flexible and allows
accurate inference and uncertainty estimates for general models in probabilistic programming
settings. We consider the computational costs of GPs specifically from
this point of view.

@@

\noindent in $D$ dimensions, where variable $\omega\in {\rm I\!R}$ denotes the frequency, and $\ell$ and $\alpha$ are the lengthscale and magnitude (marginal variance), respectively, of the kernel. The particular cases where $\nu=\infty$ and $\nu=3/2$ take the form
%
\begin{align}
S_{\infty}(\omega)&= \alpha \sqrt{2\pi}^D  \ell^D \exp(-0.5 \ell^2 \omega^2), \label{eq_specdens_inf}  \\
S_{\frac{3}{2}}(\omega)&= \alpha \, \frac{2^D\pi^{D/2}\Gamma(\frac{D+3}{2})\sqrt{3}^3}{\frac{1}{2}\sqrt{\pi}\,\ell^3}\Big(\frac{3}{\ell^2}+\omega^2 \Big)^{\!-\frac{D+3}{2}}. \label{eq_specdens_32} 
\end{align}
\todo{Is it only purpose that only some equations have numbers?}
\noindent For input dimension $D=3$ and Euclidean distance $\omega=\sqrt{\sum_{i=1}^{3}s_i^2}$, and considering a multidimensional lengthscale $\bm{\ell} \in {\rm I\!R}^{3}$, the spectral densities written above take the form
%
\begin{align*}
S_{\infty}(\omega)&= \alpha \sqrt{2\pi}^{3}  \prod_{i=1}^{3} \ell_i  \exp\!\!\big(\!-\frac{1}{2} \sum_{i=1}^{3} \ell_i^2 s_i^2 \big),   \\
S_{\frac{3}{2}}(\omega)&= \alpha \, 32\pi\sqrt{3}^3\prod_{i=1}^{3}\ell_i\big(3+\sum_{i=1}^{3}\ell_i^2 s_i^2 \big)^{\!-3}.
\end{align*}

@@

\noindent where $\beta_j \sim \text{Normal}(0,1)$. Thus, the function $f$ is approximated with a finite basis function expansion (using the eigenfunctions $\phi_j$ of the Laplace operator), scaled by the square root of spectral density values. A key property of this approximation is that the eigenfunctions $\phi_j$ do not depend on the hyperparameters of the covariance function $\theta$. Instead, the only dependence of the model on $\theta$ is through the spectral density $S_{\theta}$. The eigenvalues $\lambda_j$ are monotonically increasing with $j$ and $S_{\theta}$ goes rapidly to zero for bounded covariance functions. Therefore, eq. \eqref{eq_approxf} can be expected to be a good approximation for a finite number of $m$ terms in the series as long as the inputs values $x_i$ are not too close to the boundaries $-L$ and $L$ of $\Omega$. The computational cost of univariate HSGPs scales as $O(nm + m)$, where $n$ is the number of observations and $m$ the number of basis functions.

@@

\noindent where, again, $\beta_j \sim \text{Normal}(0,1)$. The computational cost, in learning the covariance function hyperparameters, of multivariate HSGPs scales as $O(n m^{\ast} + m^{\ast})$, where $n$ is the number of observations and $m^{\ast}$ is the number of multivariate basis functions \todo{I dont think the first of part of this sentence is correct. Do you mean the cost of evaluating the joint distribution in a single iteration rather than "the cost of learnaing the covariance function hyperparameters?}. Although this still implies linear scaling in $n$, the approximation is more costly than in the univariate case, as $m^{\ast}$ is the product of the number of univariate basis functions over the input dimensions and grows exponentially with respect to the number of  dimensions.

>>>>

\author{Gabriel Riutort-Mayol$^{1*}$, Michael R. Andersen$^2$, Paul-Christian Bürkner$^{3}$, Aki Vehtari$^{3}$}

\date{ \small
$^1$ Department of Cartographic Engineering, Geodesy, and Photogrammetry, Universitat Polit\`ecnica de Val\`encia, Spain \break
$^2$ Department of Applied Mathematics and Computer Science, Technical University of Denmark, Denmark
\break
$^3$ Department of Computer Science, Aalto University, Finland
\break
$^*$ Corresponding author, Email: gabriuma@gmail.com
}

@@

The key element of a GP is the covariance function that defines the dependence structure between function values at different inputs. However, computing the posterior distribution of a GP comes with a computational issue because of the need of inverting the covariance matrix. Given $n$ observations in the data, the computational complexity and memory requirements of computing the posterior distribution for a GP in general scale as $O(n^3)$ and $O(n^2)$, respectively. This limits their application to rather small data sets of a few tens of thousands observations at most. The problem becomes more severe when performing full Bayesian inference via sampling methods, where in each sampling step we need $O(n^3)$ computations when inverting the Gram matrix of the covariance function, usually through Cholesky factorization. To alleviate these computational demands, several approximate methods have been proposed.

@@

If $p(y_n|f(\bm{x}_n),\phi)=\mathcal{N}(y_n|f(\bm{x}_n),\sigma)$ then $\bm{f}$ can be integrated out analytically (with a computational cost of $O(n^3)$ for exact GPs and $O(nm^2)$ for sparse GPs). If $p(y_n|f(\bm{x}_n),g(\bm{x}_n))=\mathcal{N}(y_n|f(\bm{x}_n),g(\bm{x}_n))$ or $p(y_n|f(\bm{x}_n),\phi)$ is non-Gaussian, the marginalization does not have a closed form solution. Furthermore, if a prior distribution is imposed on $\phi$ and $\bm{\theta}$ to form a joint posterior for $\phi$, $\bm{\theta}$ and $\bm{f}$, approximate inference such as Markov chain Monte Carlo (MCMC) \citep{brooks_2011}, Laplace approximation (\citep{williams1998bayesian,rasmussen2006gaussian}, expectation propagation \citep{minka2001expectation}, or variational Bayes methods \citep{gibbs2000variational,csato2000efficient} are required. In this paper, we focus on the use of MCMC for integrating over the joint
posterior. MCMC is usually not the fastest approach, but it is flexible and allows
accurate inference and uncertainty estimates for general models in probabilistic programming
settings. We consider the computational costs of GPs specifically from
this point of view.

@@

\noindent in $D$ dimensions, where variable $\omega\in {\rm I\!R}$ denotes the frequency, and $\ell$ and $\alpha$ are the lengthscale and magnitude (marginal variance), respectively, of the kernel. The particular cases where $\nu=\infty$ and $\nu=3/2$ take the form
%
\begin{align}
S_{\infty}(\omega)&= \alpha \sqrt{2\pi}^D  \ell^D \exp(-0.5 \ell^2 \omega^2), \label{eq_specdens_inf}  \\
S_{\frac{3}{2}}(\omega)&= \alpha \, \frac{2^D\pi^{D/2}\Gamma(\frac{D+3}{2})\sqrt{3}^3}{\frac{1}{2}\sqrt{\pi}\,\ell^3}\Big(\frac{3}{\ell^2}+\omega^2 \Big)^{\!-\frac{D+3}{2}}. \label{eq_specdens_32} 
\end{align}
\todo{Is it only purpose that only some equations have numbers? Res: Only text referenced equations have number (Gabi). }
\noindent For input dimension $D=3$ and Euclidean distance $\omega=\sqrt{\sum_{i=1}^{3}s_i^2}$, and considering a multidimensional lengthscale $\bm{\ell} \in {\rm I\!R}^{3}$, the spectral densities written above take the form
%
\begin{align*}
S_{\infty}(\omega)&= \alpha \sqrt{2\pi}^{3}  \prod_{i=1}^{3} \ell_i  \exp\!\!\big(\!-\frac{1}{2} \sum_{i=1}^{3} \ell_i^2 s_i^2 \big),   \\
S_{\frac{3}{2}}(\omega)&= \alpha \, 32\pi\sqrt{3}^3\prod_{i=1}^{3}\ell_i\big(3+\sum_{i=1}^{3}\ell_i^2 s_i^2 \big)^{\!-3}.
\end{align*}

@@

\noindent where $\beta_j \sim \text{Normal}(0,1)$. Thus, the function $f$ is approximated with a finite basis function expansion (using the eigenfunctions $\phi_j$ of the Laplace operator), scaled by the square root of spectral density values. A key property of this approximation is that the eigenfunctions $\phi_j$ do not depend on the hyperparameters of the covariance function $\theta$. Instead, the only dependence of the model on $\theta$ is through the spectral density $S_{\theta}$. The eigenvalues $\lambda_j$ are monotonically increasing with $j$ and $S_{\theta}$ goes rapidly to zero for bounded covariance functions. Therefore, eq. \eqref{eq_approxf} can be expected to be a good approximation for a finite number of $m$ terms in the series as long as the inputs values $x_i$ are not too close to the boundaries $-L$ and $L$ of $\Omega$. The computational cost of evaluating the posterior distribution of univariate HSGPs scales as $O(nm + m)$ in a single iteration, where $n$ is the number of observations and $m$ the number of basis functions.

@@

\noindent where, again, $\beta_j \sim \text{Normal}(0,1)$. The computational cost of evaluating the posterior distribution  of multivariate HSGPs scales as $O(n m^{\ast} + m^{\ast})$ in a single iteration, where $n$ is the number of observations and $m^{\ast}$ is the number of multivariate basis functions. Although this still implies linear scaling in $n$, the approximation is more costly than in the univariate case, as $m^{\ast}$ is the product of the number of univariate basis functions over the input dimensions and grows exponentially with respect to the number of  dimensions.
